Peaks Over Threshold for Bursty Time Series

### 2018-04-24 article_springer/article_springer.Rmd
changes for submission

```{r fig:QQ-plots, fig.cap="Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\\label{fig:QQ-plots}", fig.height=5.5, fig.width=7}
par(mfrow = c(2,3))

@@

```{r solar-flare-tail-scale, message=FALSE, fig.height=4.5, fig.width=8, fig.cap="\\label{fig:flares}Stability plots for the tail and scale parameter of the Mittag-Leffler distribution of the Solar Flare dataset. Dotted horizontal lines are at $\\beta = 0.85$ and $\\sigma_0 = 3 \\times 10^7$ seconds $\\approx 0.95$ years."}
thin_flares_ctre <- flares %>% ctre() %>% thin(k = 700)
par(mfrow=c(1,2))
MLestimates(thin_flares_ctre, tail = 0.9, scale = 3 * 1E7)
```

@@

```{r flare-diagnostics-2, fig.height=7, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow = c(2,2))
empcopula(flares_ctre)
flares_ctre %>% interarrival() %>% mlqqplot(log = 'xy', tail = 0.8, main = "Mittag-Leffler QQ Plot")
```

@@

```{r hazard, fig.height=7, fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow = c(2,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

We have extended the POT (Peaks over Threshold) model, a mainstay of 
extreme value theory, to "bursty" time series, which have been studied 
intensively in statistical physics. 
Burstiness is characterized by power-law waiting times between events, 
and we have shown that the Mittag-Leffler distribution arises naturally 
as a scaling limit for the inter-exceedance times of high thresholds. 
Moreover, we have derived the following non-linear scaling behaviour: 
$\sigma \sim p^{-1/\beta}$, 
where $\sigma$ is the scale parameter of the distribution of threshold 
exceedance times, 
$p$ is the fraction of magnitudes above the threshold, 
and $\beta$ the exponent of the power law.

The "anomalous" scaling behaviour in the bursty setting entails two 
phenomena: 
i) a heavy tail of the interarrival time distribution of 
threshold crossings (long rests), and 
ii) a high propensity for more threshold crossing events immediately after 
each threshold crossing event (bursts). 
The Mittag-Leffler distribution captures both phenomena, due to its heavy
tail as well as its stretched exponential (peaked) asymptotics for small 
times. 
It generalizes the exponential distribution, and 
in the solar flare data example, this generalization 
is warranted, because the likelihood-ratio test is strongly significant.

When we introduced the CTRE model, we assumed that all events are i.i.d. 
This assumption is likely sufficient but not necessary for our limit theorem to hold. 
Morevoer, any data below a (minimum) threshold $\ell_0$ is discarded in 
our inference procedure, and hence need not satisfy the i.i.d. assumption.
For the purposes of statistical inference, we merely require that the 
inter-threshold-crossing times are i.i.d.

For some bursty time series, the current CTRE model may be too rigid, 
because often the heavy-tailed character of the inter-arrival time
distribution does not hold at all time scales; rather, it often applies at 
short and intermediate time scales, and is truncated or tempered 
(reverting to an exponential distribution) at 
very long time scales, see e.g. @MeerschaertRoyQin and @Aban06. 
In such situations, a "tempered" Mittag-Leffler distribution may provide 
a more realistic fit, which we aim to introduce in follow-up work.

>>>>

```{r QQ-plots, fig.cap="Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\\label{fig:QQ-plots}", fig.height=5.5, fig.width=7}
par(mfrow = c(2,3))

@@

```{r solar-flare-tail-scale, message=FALSE, fig.cap="\\label{fig:flares}Stability plots for the tail and scale parameter of the Mittag-Leffler distribution of the Solar Flare dataset. Dotted horizontal lines are at $\\beta = 0.85$ and $\\sigma_0 = 3 \\times 10^7$ seconds $\\approx 0.95$ years."}
thin_flares_ctre <- flares %>% ctre() %>% thin(k = 700)
par(mfrow=c(1,2))
MLestimates(thin_flares_ctre, tail = 0.9, scale = 3 * 1E7)
```

@@

```{r flare-diagnostics-2, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow=c(1,2))
empcopula(flares_ctre)
flares_ctre %>% interarrival() %>% mlqqplot(log = 'xy', tail = 0.8, main = "Mittag-Leffler QQ Plot")
```

@@

```{r hazard, fig.height=3.5 , fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow=c(1,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

We have extended the POT (Peaks over Threshold) model, a mainstay of 
extreme value theory, to "bursty" time series, which have been studied 
intensively in statistical physics. 
Burstiness is characterized by power-law waiting times between events, 
and we have shown that the Mittag-Leffler distribution arises naturally 
as a scaling limit for the inter-exceedance times of high thresholds. 
Moreover, we have derived the following non-linear scaling behaviour: 
$\sigma \sim p^{-1/\beta}$, 
where $\sigma$ is the scale parameter of the distribution of threshold 
exceedance times, 
$p$ is the fraction of magnitudes above the threshold, 
and $\beta$ the exponent of the power law. 
This "anomalous" scaling behaviour in the bursty setting entails two 
phenomena: 
i) a heavy tail of the interarrival time distribution of 
threshold crossings (long rests), and 
ii) a high propensity for more threshold crossing events immediately after 
each threshold crossing event (bursts). 
The Mittag-Leffler distribution captures both phenomena, due to its heavy
tail as well as its stretched exponential (peaked) asymptotics for small 
times. 
It generalizes the exponential distribution, and 
in the solar flare data example, this generalization 
is warranted, because the likelihood-ratio test is strongly significant.

When we introduced the CTRE model, we assumed that all events are i.i.d. 
This assumption is likely sufficient but not necessary for our limit theorem to hold. 
Moreover, any data below a (minimum) threshold $\ell_0$ is discarded for 
CTREs, and hence need not satisfy the i.i.d. assumption.
For the purposes of statistical inference, we merely require that the 
inter-threshold-crossing times are i.i.d.

The bursty CTRE approach to model "non-Poissonian" threshold crossing times
should be contrasted with the (now standard) approach of clusters of 
extremes, see e.g. @ferro2003inference. In this approach, i.i.d. event 
sequences of magnitudes are generalized to stationary sequences of
event magnitudes (subject to a mixing condition). The two approaches are 
fundamentally different: A clustering model assumes that each event 
belongs to one particular (latent) group of events. For bursts, however,
the aim is to identify an underlying scale-free pattern in the event 
dynamics, which is often characteristic of complex systems.
It is an interesting open problem to develop quality criteria, based
e.g. on measures of surprise [@Lee15], which guide an applied statistician in 
the choice between a clustering and a CTRE approach for a particular 
problem. Moreover, we believe it may be possible to unify the two approaches
by considering CTREs based on MRPs with a _stationary_, rather than 
i.i.d., sequence of magnitudes.

Finally, a purely scale-free pattern for event times may be too rigid an 
assumption for some bursty time series, 
because often the heavy-tailed character of the inter-arrival time
distribution does not hold at all time scales; rather, it applies at 
short and intermediate time scales, and is truncated (or tempered, 
reverting to an exponential distribution) at 
very long time scales [see e.g. @MeerschaertRoyQin; and @Aban06]. 
In such situations, a "tempered" Mittag-Leffler distribution may provide 
a more realistic fit, which we aim to introduce in follow-up work.

### 2018-04-24 article_springer/article_springer.tex
changes for submission

In this article, we restrict ourselves to the uncoupled case, where the
two sequences \(X(\ell, n)_{n \in \mathbb N}\) and
\(T(\ell, n)_{n \in \mathbb N}\) are independent.\footnote{To see why,
  note that \(X(\ell)\) is, in distribution, simply equal to
  \(J | J > \ell\), independent of any waiting time \(W_k\).} Figure
\ref{fig:thresholdedBursty} shows a simulated dataset in the top panel,
where \(W\) has a stable distribution with tail parameter \(\beta =\)
0.8 (and skewness \(1\) and location \(0\)), and where \(J\) is from a
standard Gumbel distribution. In the bottom panel, we plot a time series
of solar flare intensities derived from a NASA dataset (Dennis et al.
1991)\footnote{The ``complete Hard X Ray Burst Spectrometer event list''
  is a comprehensive reference for all measurements of the Hard X Ray
  Burst Spectrometer on NASA's Solar Maximum Mission from the time of
  launch on Feb 14, 1980 to the end of the mission in Dec 1989. 12,776
  events were detected, with the ``vast majority being solar flares''.
  The list includes the start time, peak time, duration, and peak rate
  of each event. We have used ``start time'' as the variable for event
  times, and ``peak rate'' as the variable for event magnitudes.}.
Clearly, the simulated data exhibit long intervals \emph{without any}
events, whereas in the real-world dataset events appear continuously.
The threshold exceedances, however, appear to have similar statistical
behaviour in both models. Observations below a threshold are commonly
discarded in Extreme Value Theory; likewise, the CTRE model interprets
these observations as noise and filters them out.

@@

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/fig:QQ-plots-1} \caption{Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\label{fig:QQ-plots}}\label{fig:fig:QQ-plots}
\end{figure}

@@

However, as described nicely by Resnick (1997), the less similar a
heavy-tailed distribution is to a Pareto distribution, the less useful a
Hill or QQ-Plot estimator becomes. The bottom-left panel of Figure
\ref{fig:QQ-plots}, for instance, shows the static QQ-estimator plot for
\(2000\) draws from the Mittag-Leffler distribution with tail parameter
\(0.8\). The stretched exponential shape means that the Mittag-Leffler
distribution has more probability near \(0\) than the Pareto
distribution, severely biasing the estimator downwards (0.62). Even by
looking at different cutoffs via the dynamic QQ-estimator plot
(bottom-center panel) one hardly identifies \(0.8\) as a clear candidate
for a tail parameter estimate.

@@

We have extended the POT (Peaks over Threshold) model, a mainstay of
extreme value theory, to ``bursty'' time series, which have been studied
intensively in statistical physics. Burstiness is characterized by
power-law waiting times between events, and we have shown that the
Mittag-Leffler distribution arises naturally as a scaling limit for the
inter-exceedance times of high thresholds. Moreover, we have derived the
following non-linear scaling behaviour: \(\sigma \sim p^{-1/\beta}\),
where \(\sigma\) is the scale parameter of the distribution of threshold
exceedance times, \(p\) is the fraction of magnitudes above the
threshold, and \(\beta\) the exponent of the power law.

The ``anomalous'' scaling behaviour in the bursty setting entails two
phenomena: i) a heavy tail of the interarrival time distribution of
threshold crossings (long rests), and ii) a high propensity for more
threshold crossing events immediately after each threshold crossing
event (bursts). The Mittag-Leffler distribution captures both phenomena,
due to its heavy tail as well as its stretched exponential (peaked)
asymptotics for small times. It generalizes the exponential
distribution, and in the solar flare data example, this generalization
is warranted, because the likelihood-ratio test is strongly significant.

When we introduced the CTRE model, we assumed that all events are i.i.d.
This assumption is likely sufficient but not necessary for our limit
theorem to hold. Moreover, any data below a (minimum) threshold
\(\ell_0\) is discarded in our inference procedure, and hence need not
satisfy the i.i.d. assumption. For the purposes of statistical
inference, we merely require that the inter-threshold-crossing times are
i.i.d. However, it might be interesting to generalize our theory to the
case of a stationary sequence of magnitudes. In case of fixed times
between observations, the distribution of the sequence of (correctly
normalized) inter-exceedance times converges under certain (mixing)
conditions to a mixture of an exponential distribution and a point mass
in zero (Ferro and Segers 2003). Hence, this also leads to a model for
extreme events that occur in cluster. Allowing for both, dependence
between the magnitudes and heavy tailed waiting times, could maybe
further improve modelling of extreme events like earthquakes, that occur
in clusters.

For some bursty time series, the current CTRE model may be too rigid,
because often the heavy-tailed character of the inter-arrival time
distribution does not hold at all time scales; rather, it often applies
at short and intermediate time scales, and is truncated or tempered
(reverting to an exponential distribution) at very long time scales, see
e.g. Mark M. Meerschaert, Roy, and Shao (2012) and Aban, Meerschaert,
and Panorska (2006). In such situations, a ``tempered'' Mittag-Leffler
distribution may provide a more realistic fit, which we aim to introduce
in follow-up work.

>>>>

In this article, we restrict ourselves to the uncoupled case, where the
two sequences \(X(\ell, n)_{n \in \mathbb N}\) and
\(T(\ell, n)_{n \in \mathbb N}\) are independent.\footnote{To see why,
  note that \(X(\ell)\) is, in distribution, simply equal to
  \(J | J > \ell\), independent of any waiting time \(W_k\).} Figure
\ref{fig:thresholdedBursty} shows a simulated dataset in the top panel,
where \(W\) has a stable distribution with tail parameter \(\beta =\)
0.8 (and skewness \(1\) and location \(0\)), and where \(J\) is from a
standard Gumbel distribution. In the bottom panel, we plot a time series
of solar flare intensities derived from a NASA dataset (Dennis et al.
1991)\footnote{The ``complete Hard X Ray Burst Spectrometer event list''
  is a comprehensive reference for all measurements of the Hard X Ray
  Burst Spectrometer on NASA's Solar Maximum Mission from the time of
  launch on Feb 14, 1980 to the end of the mission in Dec 1989. 12,776
  events were detected, with the ``vast majority being solar flares''.
  The list includes the start time, peak time, duration, and peak rate
  of each event. We have used ``start time'' as the variable for event
  times, and ``peak rate'' as the variable for event magnitudes.}.
Clearly, the simulated data exhibit long intervals \emph{without any}
events, whereas in the real-world dataset events appear continuously.
The threshold exceedances, however, appear to have similar statistical
behaviour in both models. Observations below a threshold are commonly
discarded in Extreme Value Theory (POT approach); likewise, the CTRE
model interprets these observations as noise and filters them out.

@@

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/QQ-plots-1} \caption{Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\label{fig:QQ-plots}}\label{fig:QQ-plots}
\end{figure}

@@

However, as described nicely by Resnick (1997), the less similar a
heavy-tailed distribution is to a Pareto distribution, the less useful a
Hill or QQ-Plot estimator becomes. The bottom-left panel of Figure
\ref{fig:QQ-plots}, for instance, shows the static QQ-estimator plot for
\(2000\) draws from the Mittag-Leffler distribution with tail parameter
\(0.8\). The stretched exponential shape means that the Mittag-Leffler
distribution has more probability near \(0\) than the Pareto
distribution, severely biasing the estimator downwards (0.67). Even by
looking at different cutoffs via the dynamic QQ-estimator plot
(bottom-center panel) one hardly identifies \(0.8\) as a clear candidate
for a tail parameter estimate.

@@

We have extended the POT (Peaks over Threshold) model, a mainstay of
extreme value theory, to ``bursty'' time series, which have been studied
intensively in statistical physics. Burstiness is characterized by
power-law waiting times between events, and we have shown that the
Mittag-Leffler distribution arises naturally as a scaling limit for the
inter-exceedance times of high thresholds. Moreover, we have derived the
following non-linear scaling behaviour: \(\sigma \sim p^{-1/\beta}\),
where \(\sigma\) is the scale parameter of the distribution of threshold
exceedance times, \(p\) is the fraction of magnitudes above the
threshold, and \(\beta\) the exponent of the power law. This
``anomalous'' scaling behaviour in the bursty setting entails two
phenomena: i) a heavy tail of the interarrival time distribution of
threshold crossings (long rests), and ii) a high propensity for more
threshold crossing events immediately after each threshold crossing
event (bursts). The Mittag-Leffler distribution captures both phenomena,
due to its heavy tail as well as its stretched exponential (peaked)
asymptotics for small times. It generalizes the exponential
distribution, and in the solar flare data example, this generalization
is warranted, because the likelihood-ratio test is strongly significant.

When we introduced the CTRE model, we assumed that all events are i.i.d.
This assumption is likely sufficient but not necessary for our limit
theorem to hold. Moreover, any data below a (minimum) threshold
\(\ell_0\) is discarded for CTREs, and hence need not satisfy the i.i.d.
assumption. For the purposes of statistical inference, we merely require
that the inter-threshold-crossing times are i.i.d.

The bursty CTRE approach to model ``non-Poissonian'' threshold crossing
times should be contrasted with the (now standard) approach of clusters
of extremes, see e.g. Ferro and Segers (2003). In this approach, i.i.d.
event sequences of magnitudes are generalized to stationary sequences of
event magnitudes (subject to a mixing condition). The two approaches are
fundamentally different: A clustering model assumes that each event
belongs to one particular (latent) group of events. For bursts, however,
the aim is to identify an underlying scale-free pattern in the event
dynamics, which is often characteristic of complex systems. It is an
interesting open problem to develop quality criteria, based e.g.~on
measures of surprise (Lee, Fan, and Sisson 2015), which guide an applied
statistician in the choice between a clustering and a CTRE approach for
a particular problem. Moreover, we believe it may be possible to unify
the two approaches by considering CTREs based on MRPs with a
\emph{stationary}, rather than i.i.d., sequence of magnitudes.

Finally, a purely scale-free pattern for event times may be too rigid an
assumption for some bursty time series, because often the heavy-tailed
character of the inter-arrival time distribution does not hold at all
time scales; rather, it applies at short and intermediate time scales,
and is truncated (or tempered, reverting to an exponential distribution)
at very long time scales (see e.g. Mark M. Meerschaert, Roy, and Shao
2012; and Aban, Meerschaert, and Panorska 2006). In such situations, a
``tempered'' Mittag-Leffler distribution may provide a more realistic
fit, which we aim to introduce in follow-up work.

### 2018-04-24 article_springer/article_springer_submissionfile.tex
changes for submission

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/QQ-plots-1} \caption{Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\label{fig:QQ-plots}}\label{fig:QQ-plots}
\end{figure}

@@

The authors would like to thank Prof.~Peter Scheffler for insights on
stochastic process limits for CTRMs, and Gurtek Gill who helped create
the MittagLeffleR R-package.

@@

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/simulated-example-1} \caption{Tail and scale estimates for simulated data, with waiting times drawn from the stable distribution $S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$ with $\beta = 0.8$. Dashed lines are 95\% confidence intervals, dotted lines are the known theoretical values ($0.8$ and $10000^{1/0.8}$). \label{fig:sim}}\label{fig:simulated-example}
\end{figure}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

@@

@@

\bibliographystyle{spbasic}
\bibliography{CTRMstats.bib}

>>>>

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/fig:QQ-plots-1} \caption{Pareto vs. Mittag-Leffler distribution, both with same tail exponent 0.8. Top row: solar flare magnitudes data, fitting nicely with a Pareto distribution. Bottom row: simulated Mittag-Leffler data. Left column: static QQ-Estimator plots, assessing a fit against the Pareto distribution. Centre column: dynamic QQ-Estimator plots. Right column: a usual QQ-Plot based on Mittag-Leffler quantiles.\label{fig:QQ-plots}}\label{fig:fig:QQ-plots}
\end{figure}

@@

The authors would like to thank Prof.~Peter Scheffler for insights on
stochastic process limits for CTRMs, and Gurtek Gill who helped create
the MittagLeffleR R-package.
\newpage

\section*{Appendix}

@@

\begin{figure}[b]
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/simulated-example-1} \caption{Tail and scale estimates for simulated data, with waiting times drawn from the stable distribution $S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$ with $\beta = 0.8$. Dashed lines are 95\% confidence intervals, dotted lines are the known theoretical values ($0.8$ and $10000^{1/0.8}$). \label{fig:sim}}\label{fig:simulated-example}
\end{figure}

@@

%\section*{References}\label{references}
%\addcontentsline{toc}{section}{References}
\bibliography{CTRMstats}
\bibliographystyle{spbasic}

@@

\pagebreak

@@
